Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-18 ebayes_rnaseq_heterosis.tex
changes

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhances phenotype of hybrid progeny relative to their inbred parents. The heterosis phenomenon is extensively utilized in agriculture though the molecular basis is still unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to compare expression levels of thousands of genes between parental inbred lines and their hybrid offspring to search for evidence of gene expression heterosis in RNA sequencing (RNAseq) experiments. Standard statistical approaches for RNAseq experiments have difficulty with the composite null hypothesis for gene heterosis testing. To address these shortcomings, we develop a hierarchical model based on negative-binomial data model and a computationally tractible empirical Bayes approach to evaluating gene heterosis hypotheses. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment, and then analyze that maize experiment. This article has supplementary material online. 
\end{abstract}

@@

% Need to udpate as this is directly from Tieming's paper
Heterosis, or hybrid vigor, refers to the enhances phenotype of hybrid progeny relative to their inbred parents. Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 are taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}. Since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}. Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}. Researchers speculate that gene-expression heterosis $-$ i.e, enhanced expression of one or more genes in the hybrid relative to the inbred parents $-$ could be among the mechanisms responsible for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.

Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data where the data are assumed to be continuous. They built a normal hierarchical model for gene expression that allows borrowing of information across genes for means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypothesis when testing for heterosis making many available methods ill-suited to heterosis identification. When transitioning from continuous to count data, the difficulty in a composite null hypothesis remains. We following \citeauthor{ji2014estimation} in constructing a hierarchical model now built off a negative-binomial rather than normal data model. We also utilize an empirical Bayes approach based on obtaining \emph{maximum a posterior} estimates of hyperparameters and posterior distribution for gene-specific parameters conditional on those hyperparameters.

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and calculation of posterior probabilities of gene heterosis. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating heterosis. The data and code necessary to perform this analysis is available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

An RNA sequencing (RNAseq) experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. cDNA libraries, i.e. short cDNA fragments, are constructed and next generation sequencing technology is used to determine the \emph{reads}, i.e. nucleotide sequences in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts. For more details on RNAseq experiments see \cite{nettleton2014design} and for biological background on the details of these steps see \cite{paschold2012complementation}.

To identify genes displaying heterosis using RNAseq counts, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

@@

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$. 
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g) 
\label{e:data}
\end{equation}
where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$ and $ind$ indicates the observations are conditionally independent.

A hierarchical model is constructed for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$. (These terms are the normalization factors that are typically estimated from each sample and then plugged in for further analysis.) The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independent} model where $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model where we estimate all parameters in $\Sigma$.

We assume improper uniform priors for $\eta$ and $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$ where 0 and 3 are the location and scale parameters of the untruncated distribution. In the independent model, we also $Ca^+(0,3)$ on the diagonal elements of $\Sigma$ while in the covariance model, we assume a conjugate, proper inverse Wishart with degrees of freedom $V+2$ and an identity scale matrix which results in marginally uniform correlations.

@@

The parameters from the model in Section \ref{s:model} can be group by gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$, and hyperparameter $\pi = (\gamma,\eta, \Sigma)$ where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on this estimated hyperparameter. Conditional on the hyperparameters, the gene-specific parameters are conditionally independent as in equation \eqref{e:condind}. Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
\begin{align}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right) 
\label{e:condind}
\end{align}

To perform the optimization, we use the statistical software {\tt Stan} \citep{stan-software:2014} run through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows construction of general Bayesian models with the primary restriction that all unknown parameters must be continuous. Once a model is constructed, the software can be used to obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan. We found it necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR} presumably due to the large dimension of the optimization. The {\tt Stan} software can also be used to run a Markov chain Monte Carlo (MCMC) algorithm to obtain samples from the posterior using the {\tt sampling} function in RStan. By default, Stan uses a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc} called the no-U-turn sampler \citep{hoffman2013no}.

As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate for $\theta$ and $\pi$. We explicitly use the MAP estimate for $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}. The MAP estimate for $\theta$ can also be used in this step as initial values for the MCMC.

@@

As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are three varieties of interest, i.e. $V=3$. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) which is the event of LPH or HPH. Of interest is to evaluate the hypotheses shown in equation \eqref{e:hypotheses} where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\label{e:hypotheses}
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probability, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence for extreme heterosis. This list can be used by geneticists to prioritize future experiments to understand the genetic mechanism for heterosis.

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}. Details of these data set are available in Section \ref{s:maize}. We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.

@@

Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety. We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, which resulted in 27,888 genes remaining.

To begin, we applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters. We treated these parameters as the truth for the simulation study: i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH. This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.

Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:data} independently for each gene. For each simulation, low count genes, i.e. those with three or more zeros in any variety or whose mean count across the varieties was less than one, were removed. A random set of 25,000 genes was retained from those with high enough counts. This simulation process was repeated 10 times for each of 4, 8, and 16 replicates per variety (reusing normalization factors when necessary).

@@

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}. and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.

@@

The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct. In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated using {\tt edgeR} is between those of the parents. Otherwise, the measure is the sum of the posterior probabilities of $H_2$ and $H_5$, the only hypotheses that allow for extreme heterosis.

@@

For each of the methods above as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.}
\label{f:roc}
\end{figure}
For this simulation, we can see the approaches based on the model in Section \ref{s:model}, i.e. both RStan approaches and the ShrinkBayes approach, provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases our ability to distinguish genes with heterosis from genes without improves.

Figure \ref{f:auc} provides the area under the ROC below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two. While there does not appear to be a significant difference between the independence model using {\tt RStan} or {\tt ShrinkBayes}, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.

@@

We used our method to analyze a maize data set on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17). The data are available in the Supplemental Materials. Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner NOVOALIGN. For more specific details, please see \cite{paschold2012complementation}.

A total of 39,656 genes among $V=3$ varieties each with $n_v=4$ replicates were available for analysis. As before, we eliminated any genes with three or more zeros for any variety and any genes with mean expression less than one across all samples resulting in a total of $G=27,888$ genes present in the analysis.

>>>>

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis is still unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to compare expression levels of thousands of genes between parental inbred lines and their hybrid offspring to search for evidence of gene expression heterosis in RNA sequencing (RNAseq) experiments. Standard statistical approaches for RNAseq experiments struggle with the composite null hypothesis in gene heterosis testing. To address these shortcomings, we develop a hierarchical model based on a negative-binomial data model and a computationally tractable empirical Bayes approach. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment. This article has supplementary material online.
\end{abstract}

@@

% Need to udpate as this is directly from Tieming's paper
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 grow taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}. Ever since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for the food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}. Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}. Researchers speculate that gene-expression heterosis $-$ i.e, enhanced expression of one or more genes in the hybrid relative to the inbred parents $-$ is a possible  mechanisms for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.

Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data under the assumption that these data are continuous. They built a normal hierarchical model for gene expression that allows borrowing of information across genes for means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypothesis in tests for heterosis. These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous data to count data. Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model. We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene heterosis. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating heterosis. The data and code necessary to perform this analysis is available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

@@

An RNA sequencing (RNAseq) experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. cDNA libraries, or collections of short cDNA fragments, are constructed. Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts. See \cite{nettleton2014design} for more details on RNAseq experiments, and see \cite{paschold2012complementation} for the biological background behind the above steps.

To use RNAseq counts to identify genes displaying heterosis, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

@@

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $c_{vi}$ and a gene-specific over-dispersion $\psi_g$. 
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g) 
\label{e:data}
\end{equation}
where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.

We present a hierarchical model for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$. (These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific over-dispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independent} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model in which we estimate all parameters in $\Sigma$.

We assume improper uniform priors for $\eta$ and take $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution. In the independent model, we also assign the $Ca^+(0,3)$ prior to the diagonal elements of $\Sigma$. In the covariance model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and hyperparameters $\pi = (\gamma,\eta, \Sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional of this estimated hyperparameter. Conditional on the hyperparameters, the gene-specific parameters are independent as in equation \eqref{e:condind}. Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
\begin{align}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right) 
\label{e:condind}
\end{align}

To perform the optimization, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous. With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan. It was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR}, presumably due to the large dimension of the optimization. The {\tt Stan} software also includes the RStan function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior. By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.

As we have not integrated out the gene-specific parameters when obtaining the MAP estimate, we actually receive a joint MAP estimate of $\theta$ and $\pi$. We explicitly use the MAP estimate of $\pi$ when obtaining empirical Bayes posterior samples for the gene-specific parameters as in equation \eqref{e:condind}. The MAP estimate of $\theta$ can also be used in this step as initial values for the MCMC.

@@

As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are $V=3$ varieties of interest. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of a gene in the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) to be the event of LPH or HPH. Equation \eqref{e:hypotheses} shows the most important hypotheses, where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\label{e:hypotheses}
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence of extreme heterosis. Geneticists can use this list to prioritize future experiments to understand the genetic mechanism for heterosis.

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}. Section \ref{s:maize} describes this dataset in detail. We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.

@@

Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety. We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, resulting in 27,888 remaining genes.

Next, we applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters. We treated these parameters as the truth for the simulation study: i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH. This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.

Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:data} independently for each gene. For each simulation, low count genes $-$ those with three or more zeros in any variety or whose mean count across the varieties was less than one $-$ were removed. We retained a random subset of 25,000 genes from those with high enough counts. We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.

@@

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.

@@

The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct. In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated using {\tt edgeR} is between those of the parents. Otherwise, the measure is the sum of the posterior probabilities of $H_2$ and $H_5$, the only hypotheses that allow for extreme heterosis.

@@

For each of the methods above, as well as the method described in Section \ref{s:gene_heterosis}, we sorted genes according to the plausibility of heterosis. From these sorted lists, we constructed receiver-operating characteristic (ROC) curves to evaluate the ability of these methods to identify genes with heterosis, as defined in Section \ref{s:sim_data}  \citep{landau2013dispersion}. A representative set of ROC curves is shown in Figure \ref{f:roc}. 
\begin{figure}[htbp]
\centerline{\includegraphics[width=\textwidth]{exampleROC0_1}}
\caption{Repersentative ROC curves for false positive rates below 0.1 for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}.}
\label{f:roc}
\end{figure}
For this simulation, we can see that the approaches based on the model in Section \ref{s:model} $-$ i.e. both RStan approaches and the ShrinkBayes approach $-$ provide the best true positive rate for a given false positive rate. Also, as expected, as the sample size increases, our ability to distinguish genes with heterosis from genes without improves.

Figure \ref{f:auc} provides the area under the ROC curve below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold, with the {\tt RStan} and {\tt ShrinkBayes} techniques outperforming the other two. While there does not appear to be a significant difference between {\tt ShrinkBayes} and the independence model using {\tt RStan}, performance appears to decrease for the model where a covariance is estimated, at least for smaller sample sizes.

@@

We used our method to analyze a maize data set on gene expression in parental lines B73 and Mo17 and the hybrid genotype (B73$\times$Mo17). The data are available in the Supplemental Materials. Each variety had four biological replicates measured with Illumina methodology and equipment. Reads were mapped to the whole reference genome using the short reads aligner, NOVOALIGN. For more specifics, please see \cite{paschold2012complementation}.

A total of 39,656 genes among $V=3$ varieties, each with $n_v=4$ replicates, were available for analysis. As before, we eliminated genes with three or more zeros for any variety and genes with mean expression less than one across all samples, resulting in a total of $G=27,888$ genes present in the analysis.
